---
title: "mlr3hf Hard Task: Downloading and Converting Hugging Face Datasets to mlr3 Tasks"
author: "Cao Wei"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## 1. Executive Summary and Task Objective

This live report documents the Minimum Viable Product (MVP) for the `mlr3hf` Hard Task.
The goal is to download a dataset from the Hugging Face Hub and convert it into a standard `mlr3` task using only native R tooling, without Python dependencies such as `reticulate`.

Beyond the MVP code path, this report also highlights architecture constraints, performance choices, and a practical roadmap toward a production-quality CRAN package.

## 2. Architecture Considerations and Pitfalls

Before implementation, it is useful to analyze the full data flow from API metadata to model-ready tabular data.
A robust Hugging Face interface in R must address four core concerns:

### 2.1 Lazy Loading and Out-of-Memory Risk

Large datasets should avoid eager in-memory loading.
Binary payloads should be streamed to disk whenever possible, then queried in chunks or lazily via an engine such as DuckDB.

### 2.2 Fallback Strategy and Format Compatibility

Different datasets may expose different file layouts and shard patterns.
The interface should support fallback behavior when one route fails (for example, alternate splits or multiple parquet shards).

### 2.3 Network Robustness and Custom Mirror Support

Dataset access is network-dependent.
Reliable retry logic, clear HTTP error diagnostics, and configurable endpoints are required for stable execution in diverse environments.

### 2.4 Rationale for "No Reticulate"

A pure-R stack reduces cross-language runtime complexity, improves reproducibility, and makes packaging and CI easier for CRAN workflows.

## 3. MVP Implementation: `fka/prompts.chat` Case Study

This section demonstrates an end-to-end path built with `httr2`, `duckdb`, and `data.table`.

### 3.1 Fetch Metadata and Retrieve the Parquet URL

```{r}
library(httr2)

dataset_id <- "fka/prompts.chat"
api_url <- "https://datasets-server.huggingface.co/parquet"

req <- request(api_url) |>
  req_url_query(dataset = dataset_id) |>
  req_retry(max_tries = 3)

resp <- req_perform(req)
meta <- resp_body_json(resp)

stopifnot(length(meta$parquet_files) > 0)
parquet_url <- meta$parquet_files[[1]]$url

cat("Parquet URL retrieved:\n", parquet_url, "\n")
```

### 3.2 Stream to Disk to Protect Memory

```{r}
tmp_file <- tempfile(fileext = ".parquet")

request(parquet_url) |>
  req_perform(path = tmp_file)

cat("Parquet saved to:", tmp_file, "\n")
cat("File size (bytes):", file.info(tmp_file)$size, "\n")
```

### 3.3 Query with DuckDB and Clean with data.table

```{r}
library(DBI)
library(duckdb)
library(data.table)

con <- dbConnect(duckdb())
query <- sprintf(
  "SELECT * FROM read_parquet(%s)",
  dbQuoteString(con, tmp_file)
)
data <- dbGetQuery(con, query)

setDT(data)

if ("for_devs" %in% names(data)) {
  data[, for_devs := as.factor(for_devs)]
}

if ("mlr3_row_id" %in% names(data)) {
  data[, mlr3_row_id := NULL]
}

head(data, 3)

dbDisconnect(con, shutdown = TRUE)
```

### 3.4 Construct the `mlr3` Classification Task

```{r}
library(mlr3)

stopifnot("for_devs" %in% names(data))
task <- as_task_classif(data, target = "for_devs", id = "hf_prompts_task")

task
```

## 4. Future Roadmap and API Design

For a production `mlr3hf` package, the MVP logic can be wrapped into higher-level APIs:

1. User-facing convenience wrappers such as `hftsk("fka/prompts.chat", target = "for_devs")`.
2. Automatic multi-file handling for sharded parquet datasets using DuckDB globbing (for example, `read_parquet('*.parquet')`).
3. Seamless interoperability with `mlr3pipelines` for advanced feature engineering while keeping `mlr3hf` lightweight.
4. More defensive production behavior: schema checks, type contracts, and resumable downloads.

