---
title: "mlr3hf Hard Task: Downloading and Converting Hugging Face Datasets to mlr3 Tasks"
author: "Cao Wei"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## 0. Local Environment Snapshot

Environment was detected locally on March 4, 2026.

| Item | Value |
|---|---|
| OS Kernel | Darwin 25.0.0 x86_64 |
| OS Version | macOS 26.0.1 (Build 25A362) |
| Machine Model | MacBook Pro (MacBookPro16,1) |
| CPU | Intel Core i7, 2.6 GHz, 6 cores |
| Memory | 32 GB |
| R Version | R 4.5.2 (2025-10-31) |
| C Compiler (CC) | clang -std=gnu2x |
| C++ Compiler (CXX) | clang++ -std=gnu++17 |

## 1. Executive Summary and Task Objective

This live report documents the Minimum Viable Product (MVP) for the `mlr3hf` Hard Task.
The goal is to download a dataset from the Hugging Face Hub and convert it into a standard `mlr3` task using only native R tooling, without Python dependencies such as `reticulate`.

Beyond the MVP code path, this report also highlights architecture constraints, performance choices, and a practical roadmap toward a production-quality CRAN package.


## 2. MVP Implementation: `fka/prompts.chat` Case Study

This section demonstrates an end-to-end path built with `httr2`, `duckdb`, and `data.table`.

### 3.1 Fetch Metadata and Retrieve the Parquet URL

```{r}
library(httr2)

dataset_id <- "fka/prompts.chat"
api_url <- "https://datasets-server.huggingface.co/parquet"

req <- request(api_url) |>
  req_url_query(dataset = dataset_id) |>
  req_retry(max_tries = 3)

resp <- req_perform(req)
meta <- resp_body_json(resp)

stopifnot(length(meta$parquet_files) > 0)
parquet_url <- meta$parquet_files[[1]]$url

cat("Parquet URL retrieved:\n", parquet_url, "\n")
```

### 3.2 Stream to Disk to Protect Memory

```{r}
tmp_file <- tempfile(fileext = ".parquet")

request(parquet_url) |>
  req_perform(path = tmp_file)

cat("Parquet saved to:", tmp_file, "\n")
cat("File size (bytes):", file.info(tmp_file)$size, "\n")
```

### 3.3 Query with DuckDB and Clean with data.table

```{r}
library(DBI)
library(duckdb)
library(data.table)

con <- dbConnect(duckdb())
query <- sprintf(
  "SELECT * FROM read_parquet(%s)",
  dbQuoteString(con, tmp_file)
)
data <- dbGetQuery(con, query)

setDT(data)

if ("for_devs" %in% names(data)) {
  data[, for_devs := as.factor(for_devs)]
}

if ("mlr3_row_id" %in% names(data)) {
  data[, mlr3_row_id := NULL]
}

head(data, 3)

dbDisconnect(con, shutdown = TRUE)
```

### 3.4 Construct the `mlr3` Classification Task

```{r}
library(mlr3)

stopifnot("for_devs" %in% names(data))
task <- as_task_classif(data, target = "for_devs", id = "hf_prompts_task")

task
```


## 3. Future Roadmap and API Design

### 3.1 Lessons Learned from Easy Task using `ostk`
When doing the easy task, we learned the following lessons, and they should be considered in the future design.

### 3.1.1 Lazy Loading

While reading the `mlr3oml` source code, I noticed that when `ostk()` is called, it only downloads metadata. The actual data is downloaded only when `as_task` is used or when `$data$data` is accessed. This behavior should also be integrated into our design.

### 3.1.2 Caching issue: very large datasets

Very large datasets may consist of multiple Parquet files, which differs from the MVP where only a single file is handled.
Users are also likely to want to use caching, so this needs to be handled properly.

### 3.1.3 Fallback
Has the HF data server really prepared proper Parquet files for all datasets? What if some are not ready? In that case, do we have no option, or is it necessary to download the original format (e.g., CSV) directly? Or should we just return an error?
In addition to this, setting up Parquet is a heavy operation. For example, the compiling of `duckdb` costs more than half an hour. So we need to consider the performance impact of this operation. If the user does not want to use Parquet, we should not force them to use it.

## 3.2 Robustness
At a basic level, we need robust engineering practices, such as checking HTTP status codes during network downloads. At a more advanced level, can we support resume-from-breakpoint downloads and multithreaded downloading?

## 3.3 Customized Tasks: usability-oriented interfaces

There are many advanced requirements for customized tasks, such as merging fields. Can we provide a more user-friendly interface so users do not need to manipulate `data.table` directly, manually compose view/query statements, or build `mlr3pipelines` by themselves?

## 3.4 Support conversion to multiple task types
Classification, regression, and Supervised Data Stream Classification. It seems `mlr3oml` previously could not handle task 7317 of type Supervised Data Stream Classification, and raised an error because the type was not recognized.

## 3.5 Deal with very large datasets

Very large datasets: when we eventually handle truly large datasets (e.g., 100 GB) that cannot be loaded into a `data.table` at once, we may need lower-level solutions (such as modifying the DuckDB backend dictionary mapping table `col_info`).

This is another engineering challenge caused by very large datasets. More challenges may appear later; this list only includes issues encountered so far.

## 3.6 Network constraints: user-configurable servers

For network-restricted hosts (e.g., AutoDL in China mainland) that cannot access the HF data server, we should not hard-code this URL. Users should be allowed to configure their own mirror endpoint.

## 3.7 Other considerations

We can learn from existing solutions: how do Python-based approaches solve these problems?

We need to study their source code to identify issues we may not have considered yet.

## 4. Try to do it at once [WIP]

I have already done the MVP implementation. Now with the lessons learned, why not do it at once?

